RationalGRL: A Framework for Argumentation and Goal Modeling

### 2017-01-26 background.tex
GRL section and example

\subsection{Goal-oriented Requirements Language (GRL)}
\label{sect:background:grl}
The User Requirements Notation (URN)~\cite{URN}, an ITU-T Standard, is one of the first modeling languages in the area of RE which aims at the standardization of visual notations to analyze functional (behavior) and non-functional requirements (NFRs). %such as performance, cost,  security, and usability.

URN allows software and requirements engineers to identify requirements for a proposed or an evolving system and to review such requirements for correctness and completeness. URN combines two complementary notations: the \emph{Goal-oriented Requirement Language} (GRL)~\cite{Amyot:2010:EGM:1841349.1841356} for goals and NFRs, and \emph{Use Case Maps} (UCMs)~\cite{Weiss05designing} for scenarios, business processes and functional requirements. In this article, we focus on GRL models and their links to argumentation.

GRL is a visual modeling language for specifying intentions, business goals, and non-functional requirements (NFRs) of multiple stakeholders. It can be used to specify alternatives that have to be considered, decisions that have been made, and rationales for making decisions. A GRL model is a connected graph of intentional elements that optionally are part of actors. Figure~\ref{fig:SchedulerModel} illustrates a GRL diagram. An actor (\includegraphics[scale=1]{img/actor}) represents a stakeholder of a system, or the system itself (e.g. In Figure~\ref{fig:SchedulerModel}, \textsf{Meeting Participant}). Actors are holders of intentions; they are the active entities in the system or its environment who want goals to be achieved, tasks to be performed, resources to be available, and softgoals to be satisfied. Softgoals (\includegraphics[scale=1]{img/softgoal}) differentiate themselves from goals (\includegraphics[scale=1]{img/goal}) in that there is no clear, objective measure of satisfaction for a softgoal whereas a goal is quantifiable, often in a binary way. Softgoals (e.g. \textsf{User Friendly)} are often more related to NFRs, whereas goals (such as \textsf{Agreeable Meeting Dates}) are more related to functional requirements. Tasks (\includegraphics[scale=1]{img/task}) represent solutions to (or operationalizations of) goals and softgoals. In Figure~\ref{fig:SchedulerModel}, some of the tasks are \textsf{Attend Meetings} and \textsf{Arrange Meetings}. In order to be achieved or completed, softgoals, goals, and tasks may require resources (\includegraphics[scale=1]{img/resource}) to be available (e.g. \textsf{Agreement})

\begin{figure*}[ht]
\centering
\includegraphics[scale=0.35]{img/SchedulerModel}
\caption{GRL Model for Online Meeting Scheduler}
\label{fig:SchedulerModel}
\end{figure*}

Different links connect the elements in a GRL model. AND, IOR, and XOR decomposition links (\includegraphics[scale=1]{img/decomposition}) allow an element to be decomposed into sub-elements. In Figure~\ref{fig:SchedulerModel}, the task \textsf{Participate in Meetings} is AND-decomposed to the tasks \textsf{Attend Meetings} and \textsf{Arrange Meetings} as well as the softgoal \textsf{Convenient Meeting Date}. Contribution links (\includegraphics[scale=1]{img/contribution}) indicate desired impacts of one element on another element. A contribution link has a qualitative contribution type or a quantitative contribution. Softgoal \textsf{User Friendly} has \textsf{some positive} qualitative contribution to the softgoal \textsf{Low Efforts}. Dependency links (\includegraphics[scale=1]{img/dependency}) model relationships between actors. For example, actor \textsf{Meeting Scheduler} depends on the actor \textsf{Meeting Participant} to \textsf{Agree to Date} via the resource \textsf{Proposal Dates} to fulfill its task \textsf{Schedule a Meeting}.

GRL is based on $i*$~\cite{Yu:1997:TMR:827255.827807} and the NFR Framework~\cite{chung2012non}, but it is not as restrictive as $i*$. Intentional elements and links can be more freely combined, the notion of agents is replaced with the more general notion of actors, i.e., stakeholders, and a task does not necessarily have to be an activity performed by an actor, but may also describe properties of a solution.

\todo{Floris}{Sepideh}{can you haver a look at the motivation for using GRL (below)? It used to be a separate section but I think it fits best here, have a look at what we need. I agree that we can mention jUCMNav but not use it as motivation since we will use online tool.}

GRL is an international standard with well defined syntax and semantics. This allows us to combine the theory of formal argumentation with GRL precisely, hereby satisfying requirement 3 as described in the introduction. Moreover, GRL has an open source tool called jUCMNav, which simplifies an implementation. Since implementation is a key concern for us, the extensive tool support is a strong argument in favor of GRL.

According to Amyot et al.~\cite{amyot2009lightweight}, GRL has several benefits over other existing goal modeling languages, such as integration with a scenario notation (UCM), support for both qualitative and quantitative attributes (for contributions levels and satisfaction levels), there is a clear separation of GRL model elements from their graphical representation, and it provides support for providing a scalable and consistent representation of multiple views/diagrams of the same goal model (see~\cite[Ch.2]{Ghanavati2013} for an more details). Finally, URN and jUCMNav provide traceability links between concepts and instances of goal modeling and behavioral design models. This traceability enables direct impact analysis of goal changes on a design and vice versa. This traceability feature seems to be a good fit with our current research: We aim to add traceability links between intentional elements and their underlying arguments. \todo{Sepideh}{Sepideh}{I need to improve: GRL has has the capability to be extended via profile, metadata and rules, and due to this capability, it has been extended to many domain. This feature also helps us to apply our argumentation to other domain such as compliance or enterprise architecture. }

\subsubsection{Online Meeting Scheduler Example}
\label{sect:background:casestudy}

@@

The GRL model in Figure~\ref{fig:SchedulerModel} shows the softgoals, goals, tasks and the relationship between the different intentional elements in the model. However, the rationales and arguments behind certain intentional elements have not been discussed or illustrated in the GRL model. Some of the questions that might be interesting to know about are the following:

@@

\todo[resolved]{Sepideh}{all}{We need to further improve this part}
\todo{Marc}{all}{Improved this part, what do you think?}
These are the type of the questions that we cannot answer just by looking at the GRL models. The model in figure~\ref{fig:SchedulerModel} does not contain information about discussions that let up to the resulting elements of the model, such as by various clarification steps for the naming, or alternatives that have been considered for the relationships. In order to address this, we use Argument Schemes for Goal Modeling (GMAS). GMAS can help us decide what intentional elements should remain in GRL and what new ones to be added or deleted, provide rationale behind the design decisions and the relationships between the links.

@@

These critical questions can point to new arguments that might counter the original argument. Take, for example, critical question 8. If new evidence comes up indicating that meeting schedulers do not care about user friendliness we have a counterargument to the above argument, `evidence shows that \textsf{user friendliness} is not a legitimate value'. Another way to counter an argument for an action is to suggest an alternative action that realizes the same goal (question 2) or an alternative goal that promotes the same value (question 3). For example, we can argue that \textbf{Find agreeable date using scheduler} also realizes the goal \textbf{Find agreeable meeting date}, which gives us a counterargument to the original argument -- to find a date by talking to the initiator -- that also follows PRAS.

>>>>

\subsection{Goal-oriented Requirements Language (GRL)}
\label{sect:background:grl}
GRL is a visual modeling language for specifying intentions, business goals, and non-functional requirements (NFRs) of multiple stakeholders \cite{Amyot:2010:EGM:1841349.1841356}. It can be used to specify alternatives that have to be considered, decisions that have been made, and rationales for making decisions. A GRL model is a connected graph of intentional elements that optionally are part of actors. Figure~\ref{fig:trafficsim} illustrates a GRL diagram. An actor (\includegraphics[scale=1]{img/actor}) represents a stakeholder of a system (\textbf{User}, Figure~\ref{fig:trafficsim}), or the system itself (\textbf{Simulation}, Figure~\ref{fig:trafficsim}). Actors are holders of intentions; they are the active entities in the system or its environment who want goals to be achieved, tasks to be performed, resources to be available, and softgoals to be satisfied. Softgoals (\includegraphics[scale=1]{img/softgoal}) differentiate themselves from goals (\includegraphics[scale=1]{img/goal}) in that there is no clear, objective measure of satisfaction for a softgoal whereas a goal is quantifiable, often in a binary way. Softgoals (e.g. \textbf{Realistic simulation)} are often more related to NFRs, whereas goals (such as \textbf{Generate cars}) are more related to functional requirements. Tasks (\includegraphics[scale=1]{img/task}) represent solutions to (or operationalizations of) goals and softgoals. In Figure~\ref{fig:trafficsim}, some of the tasks are \textbf{Create new cars} and \textbf{Keep same cars}. In order to be achieved or completed, softgoals, goals, and tasks may require resources (\includegraphics[scale=1]{img/resource}) to be available (e.g. \textbf{******})

\begin{figure*}[ht]
\centering
\includegraphics[scale=0.35]{img/Example}
\caption{GRL Model for the traffic simulator}
\label{fig:trafficsim}
\end{figure*}

Different links connect the elements in a GRL model. AND, IOR, and XOR decomposition links (\includegraphics[scale=1]{img/decomposition}) allow an element to be decomposed into sub-elements. In Figure~\ref{fig:trafficsim}, the goal \textbf{Generate cars} is XOR-decomposed to the tasks \textbf{Create new cars} and \textbf{Keep same cars}. Contribution links (\includegraphics[scale=1]{img/contribution}) indicate desired impacts of one element on another element. A contribution link has a qualitative contribution type or a quantitative contribution. Task \textbf{Create new cars} has a \textbf{help} qualitative contribution to the softgoal \textbf{Realistic simulation}. Dependency links (\includegraphics[scale=1]{img/dependency}) model relationships between actors. For example, actor \textbf{Simulation} depends on the actor \textbf{User} to perform the task \textbf{Adjust car spawning rate} to fulfil its task \textbf{Create new cars}.

GRL is based on $i*$~\cite{Yu:1997:TMR:827255.827807} and the NFR Framework~\cite{chung2012non}, but it is not as restrictive as $i*$. Intentional elements and links can be more freely combined, the notion of agents is replaced with the more general notion of actors, i.e., stakeholders, and a task does not necessarily have to be an activity performed by an actor, but may also describe properties of a solution. GRL has a well-defined syntax and semantics, which are necessary if we want to incorporate it into a formal framework (requirements 1 and 2 as described in the introduction). Furthermore,GRL provides support for providing a scalable and consistent representation of multiple views/diagrams of the same goal model (see~\cite[Ch.2]{Ghanavati2013} for an more details). For example, traceability links are provided between between concepts and instances of goal modeling and behavioral design models \cite{}. Multiple views and traceability are a good fit with our current research: we aim to add traceability links between intentional elements and their underlying arguments. \todo{Sepideh}{Sepideh}{I need to improve: GRL has has the capability to be extended via profile, metadata and rules, and due to this capability, it has been extended to many domain. This feature also helps us to apply our argumentation to other domain such as compliance or enterprise architecture. }

\subsubsection{Traffic simulator example}
\label{sect:background:casestudy}

@@

The GRL model in Figure~\ref{fig:trafficsim} shows the softgoals, goals, tasks and the relationship between the different intentional elements in the model. However, the rationales and arguments behind certain intentional elements have not been discussed or illustrated in the GRL model. Some of the questions that might be interesting to know about are the following:

@@

\todo[resolved]{Sepideh}{all}{We need to further improve this part}
\todo{Marc}{all}{Improved this part, what do you think?}
These are the type of the questions that we cannot answer just by looking at the GRL models. The model in figure~\ref{fig:trafficsim} does not contain information about discussions that let up to the resulting elements of the model, such as by various clarification steps for the naming, or alternatives that have been considered for the relationships. In order to address this, we use Argument Schemes for Goal Modeling (GMAS). GMAS can help us decide what intentional elements should remain in GRL and what new ones to be added or deleted, provide rationale behind the design decisions and the relationships between the links.

@@

These critical questions can point to new arguments that might counter the original argument. Take, for example, critical question 8. If new evidence comes up indicating that meeting schedulers do not care about user friendliness we have a counterargument to the above argument, `evidence shows that \textbf{user friendliness} is not a legitimate value'. Another way to counter an argument for an action is to suggest an alternative action that realizes the same goal (question 2) or an alternative goal that promotes the same value (question 3). For example, we can argue that \textbf{Find agreeable date using scheduler} also realizes the goal \textbf{Find agreeable meeting date}, which gives us a counterargument to the original argument -- to find a date by talking to the initiator -- that also follows PRAS.

### 2017-01-26 introduction.tex
GRL section and example

The aim of this research is to resolve the above issues by designing a framework, which combines a methodology for the goal modeling process \todo{Floris}{Sepideh}{Do you know of any generic methodologies or heuristics for goal modeling (akin to Juretas GAM)?} with ideas from Artificial Intelligence on practical reasoning and argumentation \cite{atkinson2007} and simple software tool support. We identified several important requirements for our framework: (1) it must be able to formally model parts of the discussion process in the early-requirements phase of an information system, (2) it must be able to generate goal models based on the discussions, (3) it should have formal traceability links between goal elements and arguments in the discussions, (4) it must have tool support, (5) a methodology must be identified that allows the framework to be used by practitioners, and (6) the framework must identify arguments and rationales that might not have been found, or might have been lost, otherwise \todo{Floris}{all}{I think (6) should be replaced with something we actually do in the paper: we don't show that arguments etc. allows us to find rationales that would otherwise not have been found -- for this we would need an experiment. Tailor it towards our empirical work: ``the framework should be close to on the actual discussions stakeholders or designers have in the early requirements engineering phase''}.

>>>>

The aim of this research is to resolve the above issues by designing a framework, which combines a methodology for the goal modeling process \todo{Floris}{Sepideh}{Do you know of any generic methodologies or heuristics for goal modeling (akin to Juretas GAM)?} with ideas from Artificial Intelligence on practical reasoning and argumentation \cite{atkinson2007} and simple software tool support. We identified several important requirements for our framework: (1) it must be able to formally model parts of the discussion process in the early-requirements phase of an information system; (2) it must be able to generate formal goal models based on the discussions; (3) it should have formal traceability links between goal elements and arguments in the discussions; (4) it must have tool support; (5) a methodology must be identified that allows the framework to be used by practitioners; and (6) the framework must identify arguments and rationales that might not have been found, or might have been lost, otherwise \todo{Floris}{all}{I think (6) should be replaced with something we actually do in the paper: we don't show that arguments etc. allows us to find rationales that would otherwise not have been found -- for this we would need an experiment. Tailor it towards our empirical work: ``the framework should be close to on the actual discussions stakeholders or designers have in the early requirements engineering phase''}.
